Extended Data

### 2022-02-24 doc/manuscript/Dow_manuscript_Nature.Rmd
remove DOY_g_max from main text
#119

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$) and the DOY on which it occurred ($DOY_{g_{max}}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- (67 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (52 trees at SCBI and 183 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022]. 
These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of seasonal milestones.
Consistent with findings from previous studies[@parmesan_globally_2003;@ibanez_forecasting_2010;@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
Similarly, with the exception of ring-porous $DOY_{75}$ at SCBI and ring-porous $DOY_{g_{max}}$ at HF, the stem growth milestones $DOY_{25}$, $DOY_{50}$, $DOY_{75}$,  and $DOY_{g_{max}}$ decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 0-1.7 days/ $^\circ$C for ring-porous species and 2.8-2.9 days/ $^\circ$C for diffuse-porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring-porous species and 0.9-4.2 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise failed to translate extended growing seasons into biologically significant increases in stem growth, and thereby have negligible effect on total annual growth for most species and locations (Fig. 3).**
**We demonstrate that warmer springs either have little effect on, or even hasten, the deceleration of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`).** <!--repeated sentence/ CD made potential changes to the first one. See what you think. We should mention how 1999 was a drought year and also the warmest year in the CTW at HF for both wood types. This is potentially contributing to the negative correlations between CTW TMAX and total growth at HF. Maybe if there wasn't a severe drought in the warmest year we would see positive correlations like we do at SCBI. This is also probably worth mentioning when we discuss the influence of climate during peak growing season being greater than the spring climate. -->
Our observations suggest that the deceleration of stem expansion (starting on $DOY_{g_{max}}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016]. <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.--> <!--Krista, Also note that maxrate_DOY was advanced in RP+DP at SCBI but only DP at HF. Maxrate_DOY for RP at HF was actually significantly delayed. This could be a consequence of the severe drought in 1999 which we've included in our analysis. My brain is completely fried to the point that I barely remember what the literature claims this max rate is related to (water stress, right?). So Maybe you're in a better position to add some discussion about this here. -->
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

@@

We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define stem growth milestones and growth rates (Fig. `r schematic_summary`). 
In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as: 
$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
Here, $L$ and $K$ are lower and upper asymptotes of the model. 
$DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
$r$ shapes the slope of the curve at the inflection point,
and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
This allows an asymmetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth.
The DOY on which maximum growth occurs, $DOY_{g_{max}}$ (Fig. `r schematic_summary`), occurs on $DOY_{ip}$ when $\theta = 1$.
The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year. 
We used the first and last manual dendrometer band measurements as starting points to calculate these two variables. 
With this method, initial measurements too late in the growing season which miss the start of the growing season can cause incorrect estimations of $a$ and $b$. 
The 1998 survey at Harvard Forest recorded first measurements on May 26, which is less than 2 weeks away from the average DOY25 at this site. 
This late start made the calculation of timing and total growth variables unreliable, so we removed tree years in 1998 from our analysis. 
The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015]. 
These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year. 
An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015;@herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest. 
Having fit the model for each band-year of data, we then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).

@@

<!--Hypothesis testing -->
Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $DOY_{g_{max}}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig. `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the CTW, along with random effects of species and of individual tree. 
We ran separate models for each site, and for the response of all growth parameters to $T_{max}$ (or $T_{min}$). 
This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@stan_development_team_stan_2019]. 
In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
More information on rstanarm's default weakly informative priors:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->

>>>>

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), the maximum daily growth rates ($g_{max}$; occurs on average within 2 days of $DOY_{50}$), and total annual increment in diameter at breast height ($\Delta DBH$; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- (67 trees at SCBI and 138 trees at Harvard Forest) and diffuse-porous (52 trees at SCBI and 183 trees at Harvard Forest) species, which differ in the seasonal timing of growth[@dorangeville_peak_2021;@tumajer_limitation_2022]. 
These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

For ring- and diffuse- porous species at both sites, warmer $T_{max}$ in the CTW resulted in earlier achievement of seasonal milestones.
Consistent with findings from previous studies[@parmesan_globally_2003;@ibanez_forecasting_2010;@friedl_tale_2014], leaf phenological milestones advanced at both sites (Fig. `r DOY_timing`a-b, `r EDT_phenology_parameters`), with greenup (DOY when EVI2 first crossed 15% of the segment EVI2 amplitude) advancing 3.45 days/$^\circ$C at SCBI (p=0.016) and 2.4 days/ $^\circ$C at Harvard Forest (p=0.1).
Similarly, with the exception of ring-porous $DOY_{75}$ at SCBI, the stem growth milestones $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ decreased with mean $T_{max}$ during the critical temperature window (Figs. `r schematic_summary`, `r DOY_timing`c-d; `r EDF_pheno_all`).
Specifically, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ advanced 0-1.7 days/ $^\circ$C for ring-porous species and 2.8-2.9 days/ $^\circ$C for diffuse-porous species at SCBI, and 10.3-12.3 days/ $^\circ$C for ring-porous species and 0.9-4.2 days/ $^\circ$C for diffuse-porous species at Harvard Forest (`r EDT_phenology_parameters`).

@@

Together, our results demonstrate that warmer spring temperatures in the temperate deciduous forests of eastern North America advance the timing of stem growth but have little effect on total annual growth (Figs. `r schematic_summary`- `r tree_rings`). 
The observed advance in stem growth under warmer springs parallels phenological advances observed for canopy foliage (Fig. `r DOY_timing`a-b)[@keenan_net_2014;@finzi_carbon_2020;@stuble_plant_2021], which makes sense given close coordination of budburst and xylogenesis[@delpierre_temperate_2016], and _NEE_[@keenan_net_2014;@finzi_carbon_2020]. 
**However, inconsistent with the concept that an earlier start to growth would increase annual woody productivity, we demonstrate that warmer springs either hasten the deceleration of stem expansion or otherwise failed to translate extended growing seasons into biologically significant increases in stem growth, and thereby have negligible effect on total annual growth for most species and locations (Fig. 3).**
**We demonstrate that warmer springs either have little effect on, or even hasten, the deceleration of stem expansion and thereby have negligible effect on total annual stem growth for most species and locations (Fig. `r tree_rings`).** <!--repeated sentence/ CD made potential changes to the first one. See what you think. We should mention how 1999 was a drought year and also the warmest year in the CTW at HF for both wood types. This is potentially contributing to the negative correlations between CTW TMAX and total growth at HF. Maybe if there wasn't a severe drought in the warmest year we would see positive correlations like we do at SCBI. This is also probably worth mentioning when we discuss the influence of climate during peak growing season being greater than the spring climate. -->
Our observations suggest that the deceleration of stem expansion (starting around $DOY_{50}$), which occurs mid-summer near the time of <!--the summer solstice and--> peak canopy greenness (`r EDF_growth_curves_all`)[@finzi_carbon_2020;@etzold_number_2022], is likely driven by cues other than photosynthate limitation, such as water stress, daylength, or sink limitation[@zweifel_why_2021;@etzold_number_2022;@zani_increased_2020;@keenan_timing_2015;@fu_variation_2014;@delpierre_wood_2016;@delpierre_temperate_2016]. <!--note that the timing of the cessation of PRIMARY shoot growth is linked to bud-set, which must be accomplished before the first frost, and has been the subject of some research (see delpierre_temperate_2016, rohde_temperature_2011), but I don't know what if any linkage exists between secondary growth and bud set.--> <!--Krista, Also note that maxrate_DOY was advanced in RP+DP at SCBI but only DP at HF. Maxrate_DOY for RP at HF was actually significantly delayed. This could be a consequence of the severe drought in 1999 which we've included in our analysis. My brain is completely fried to the point that I barely remember what the literature claims this max rate is related to (water stress, right?). So Maybe you're in a better position to add some discussion about this here. -->
Our tree-ring analysis (Fig. `r tree_rings`) demonstrates that the primary effect of warming temperatures on annual ring-width is not an augmentation through an earlier start to growth, but rather a reduction associated to drought stress during the peak growing season[@dorangeville_drought_2018;@etzold_number_2022]. 
Warm springs may also amplify summer drought stress in some times and places, effectively canceling out any positive effects of an extended growing period[@buermann_widespread_2018;@zhang_drought_2021]. 
However, we did not find evidence that higher spring temperatures were correlated with summer drought stress in our dendrometer band analysis, implying that the effects of warm spring temperatures on stem growth elucidated here (Fig. `r schematic_summary`) were not attributable to summer drought.

@@

We fit a five-parameter logistic growth model[@mcmahon_general_2015] to dendrometer band data from each tree-year to define stem growth milestones and growth rates (Fig. `r schematic_summary`). 
In particular, we model the observed diameter at breast height (DBH) on a given day of the year (DOY; *i.e.*, julian days) as: 
$$ DBH = L + \frac{K - L} { 1 + 1/\theta \cdot exp[-r(DOY - DOY_{ip}) / \theta)] ^ \theta}$$ 
Here, $L$ and $K$ are lower and upper asymptotes of the model. 
$DOY_{ip}$ is the day of year where the inflection point in growth rate occurs,
$r$ shapes the slope of the curve at the inflection point,
and $\theta$ is a tuning parameter controlling the slope of the curve toward the upper asymptote.
This allows an asymmetric fit to the data, where the onset of growth can be estimated independent of the cessation of growth. 
When $\theta = 1$, $g_{max}$ occurs on $DOY_{ip}$.
The model outputs two additional variables, $a$ and $b$, which represent the beginning and end DBH in each model year. 
We used the first and last manual dendrometer band measurements as starting points to calculate these two variables. 
With this method, initial measurements too late in the growing season which miss the start of the growing season can cause incorrect estimations of $a$ and $b$. 
The 1998 survey at Harvard Forest recorded first measurements on May 26, which is less than 2 weeks away from the average DOY25 at this site. 
This late start made the calculation of timing and total growth variables unreliable, so we removed tree years in 1998 from our analysis. 
The model was fit in R v4.0 using the functions developed in the *Rdendrom* package[@mcmahon_general_2015]. 
These functions take the time-series of manual dendrometer band measurements and return maximum-likelihood optimized values of the above five parameters that best predict DBH for each day of year. 
An advantage of this approach is that short-term shrinkage and swelling associated with rain events [@mcmahon_general_2015;@herrmann_tree_2016] and measurement errors show up as residual variation and do not unduly influence the parameters of interest. 
Having fit the model for each band-year of data, we then modeled DBH using these optimal parameter values in our logistic growth model and extracted the intra-annual growth variables of interest (Fig. `r schematic_summary`).

@@

<!--Hypothesis testing -->
Correlation between the dendrometer band-derived growth parameters ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, $L_{PGS}$, $g_{max}$, and $\Delta DBH$}, Fig. `r schematic_summary`) and spring temperatures were assessed using a linear mixed model in a hierarchical Bayesian framework.
Analyses were run for both $T_{max}$ and $T_{min}$, with qualitatively similar results, but we present only results for $T_{max}$, which had overall stronger correlation with growth parameters. 
Mixed effects models were used to test the response of growth parameters to fixed effects of xylem porosity and mean $T_{max}$ (or $T_{min}$) during the CTW, along with random effects of species and of individual tree. 
We ran separate models for each site, and for the response of all growth parameters to $T_{max}$ (or $T_{min}$). 
This mixed-effect model was run within a hierarchical Bayesian framework and fit using the `rstanarm` R interface to the Stan programming language[@gabry_rstanarm_2020;@stan_development_team_stan_2019]. 
In all cases unless otherwise specified, all prior distributions are set to be the weakly informative defaults. 
<!--
More information on rstanarm's default weakly informative priors:
https://cran.r-project.org/web/packages/rstanarm/vignettes/priors.html#default-weakly-informative-prior-distributions
-->
